### 2014-11-13 implementation.tex
Fix verbatim issues.
Fixed more typos.
Fixed several typos.

The implementation of LM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.
The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
Here, we review the most important details of that paper, while focusing on the
changes for efficiently implementing coordination.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible for
executing a subset of \emph{active nodes}. A node is active if it has unexamined
facts. After a node is processed, it becomes inactive until a new fact is
derived for it. When a new fact is derived for an
inactive node, the node is made active and placed on the appropriate queue of
the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread.
If unsuccessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

\paragraph{Nodes} 
A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available facts and potential candidate rules. We need
a separate indexing structure per node since rules run local to each node.

@@

\begin{tightdescription}
   \item[working]: the node is executing.
   \item[inactive]: the node is inactive, i.e., it has no new facts and is not in any
   queue for processing.
   \item[queue]: the node is active with new facts and is in some queue waiting
   to be processed.
   \item[stealing]: the node has just been stolen and is in the process of being
   moved to another thread.
   \item[coordinating]: the node is moving from one queue to another.
\end{tightdescription}

@@

Each node is protected by a main spin-lock that allows threads to
change node attributes: incoming facts, owner thread, state variable
and locality information.  A database spin-lock is locked whenever the
node enters into the \textbf{working} state.

@@

To support priorities each thread has two pairs of queues: a pair of
doubly linked lists known as the \emph{standard queue} and a pair of
\emph{min/max} heap known as the \emph{priority queue}.  The standard queue
contains nodes without priorities and supports push into tail, remove
node from the head, remove arbitrary node, and remove first half of
nodes.  The priority queue contains nodes with priorities and is
implemented as a binary heap array. It supports the following
operations: push into the heap, remove the \emph{min} node, remove an
arbitrary node, remove half of the nodes (horizontal split), and
priority update.  Operations for removing half of the queue are
implemented in order to support node stealing, while operations to
remove arbitrary nodes or update the priority of nodes allows threads
to change the priority of nodes.

@@

The \mytt{next} and \mytt{prev} pointers of the regular queue are
part of the node structure in order to save space. These pointers are
also used as the index in the priority queue and current priority,
respectively.

Both the regular and priority queue are implemented as a pair of
queues.  This first of the pair is the \emph{static queue} which
contains nodes that cannot be stolen.  The other queue in each pair is
the \emph{dynamic queue} which contains nodes which can be stolen by
other threads.

@@

Threads synchronize with each other using mutual exclusion. We use a spin-lock in
each queue to protect queue operations.
Given threads $T_1$ and $T_2$, we enumerate the most important
synchronization intensive places in the virtual machine:

\begin{tightdescription}
   \item[New facts:] When a node executes in $T_1$ and derives facts
   to a node in $T_2$, $T_1$ first buffers the facts 
   and then sends them to the target node. Here, it checks if the
   node is currently \textbf{idle} and then synchronizes with $T_2$ to add the
   node to the $T_2$'s queue.
   \item[Thread activation:] If $T_2$ is inactive when adding facts to a node in
   $T_2$, $T_1$ also synchronizes with $T_2$ to change $T_2$'s state to \emph{active}.
   \item[Node stealing:] $T_1$ synchronizes with $T_2$ when it attempts to steal
   nodes from $T_2$ by removing half of the nodes from one of $T_2$'s queues.
   \item[Coordination:] If $T_1$ needs to perform coordination operations
   to a node in $T_2$, it may need to synchronize with $T_2$ during priority
   updates in order to move the node in $T_2$'s queues.
\end{tightdescription}

@@

In the paper~\cite{cruz-ppdp14}, LM programs were compared to similar
programs written in other languages. The Loopy Belief Propagation
program was around 2 times slower than GraphLab, while the N Queens
program was 10 to 15x slower than a sequential C implementation. The
LM programs had similar performance to programs written in Python.
The introduction of coordination in CLM has little or no impact on the
overall performance of these programs.  In fact, due to some
improvements in the virtual machine, programs such as PageRank and
Greedy Graph Coloring now perform slightly better than before.

>>>>

The implementation of CLM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.
The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
Here, we review the most important details of that paper, while focusing on the
changes for efficiently implementing coordination.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible for
executing a subset of \emph{active nodes}. A node is active if it has unexamined
facts. After a node is processed, it becomes inactive until a new fact is
derived for it. When a new fact is derived for an
inactive node, the node is made active and placed on the appropriate queue of
the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
thread does not have nodes to process, it attempts to steal nodes from a random thread.
If unsuccessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

\paragraph{Nodes} 
A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available facts and potential candidate rules. We need
a separate indexing structure per node since rules are locally run.

@@

\begin{tightitemize}
   \item \textbf{working}: the node is executing.
   \item \textbf{inactive}: the node is inactive, i.e., it has no new facts and is not in any
   queue for processing.
   \item \textbf{queue}: the node is active with new facts and is in some queue waiting
   to be processed.
   \item \textbf{stealing}: the node has just been stolen and is in the process of being
   moved to another thread.
   \item \textbf{coordinating}: the node is moving from one queue to another.
\end{tightitemize}

@@

Each node is protected by a main spin-lock that allows threads to
change node attributes: incoming facts, owner thread, state variable
and locality information.  A database spin-lock is locked when the
node is in the \textbf{working} state.

@@

To support priorities each thread has two pairs of queues: a pair of
doubly linked lists known as the \emph{standard queue} and a pair of
\emph{min/max} heaps known as the \emph{priority queue}.  The standard queue
contains nodes without priorities and supports push into tail, remove
node from the head, remove arbitrary node, and remove first half of
nodes.  The priority queue contains nodes with priorities and is
implemented as a binary heap array. It supports the following
operations: push into the heap, remove the \emph{min} node, remove an
arbitrary node, remove half of the nodes (horizontal split), and
priority update.  Operations for removing half of the queue are
implemented in order to support node stealing, while operations to
remove arbitrary nodes or update priority allows threads
to change the priority of nodes.

@@

The \mytt{next} and \mytt{prev} pointers of the regular queue are
part of the node structure in order to save space. These pointers are
also used as the index in the priority queue and current priority,
respectively. Both the regular and priority queue are implemented as a pair of
queues.  This first queue is the \emph{static queue} which
contains nodes that cannot be stolen.  The other queue is
the \emph{dynamic queue} which contains nodes which can be stolen by
other threads.

@@

Threads synchronize with each other using mutual exclusion. We use a spin-lock in
each queue to protect queue operations.
Given threads $T_1$ and $T_2$, we enumerate the synchronization hotspots
in the virtual machine:

\begin{tightitemize}
   \item \textbf{New facts}: When a node executes in $T_1$ and derives facts
   to a node in $T_2$, $T_1$ first buffers the facts 
   and then sends them to the target node. Here, it checks if the
   node is currently \textbf{idle} and then synchronizes with $T_2$ to add the
   node to the $T_2$'s queue.
   \item \textbf{Thread activation}: If $T_2$ is inactive when adding facts to a node in
   $T_2$, $T_1$ also synchronizes with $T_2$ to change $T_2$'s state to \emph{active}.
   \item \textbf{Node stealing}: $T_1$ synchronizes with $T_2$ when it attempts to steal
   nodes from $T_2$ by removing half of the nodes from one of $T_2$'s queues.
   \item \textbf{Coordination}: If $T_1$ needs to perform coordination operations
   to a node in $T_2$, it may need to synchronize with $T_2$ during priority
   updates in order to move the node in $T_2$'s queues.
\end{tightitemize}

@@

In the paper~\cite{cruz-ppdp14}, LM programs were compared to similar
programs written in other languages. The Loopy Belief Propagation
program was around 2 times slower than GraphLab, while the N Queens
program was 10 to 15x slower than a sequential C implementation.
Furthermore, LM programs had better performance than programs written in Python.
We measured the impact of our implementation changes on unmodified LM programs
and our experiments indicate that there is little to no impact on the overall
performance of these programs. In fact, due to some
improvements in the virtual machine, programs such as PageRank and
Greedy Graph Coloring now perform slightly better than before.

### 2014-11-13 locks.tex
Fixed several typos.

As can be seen in Table~\ref{queue_instructions} all the coordinated
versions of the programs require significantly more queue operations.
The data for this table comes from recording the number of queue
operations and facts derived.  Queue operations represent the number
of normal queue operations executed (each costs 1) plus the number of
\emph{percolate-up}/\emph{percolate-down} operations executed for each
manipulation of the priority queue.  As expected, programs which rely
on prioritizing the order of execution to obtain better asymptotic
complexity (e.g., SSSP) improve in overall running time at the cost of
more queue operations.  In all cases, coordination adds a significant
number of total queue operations, but the resulting overhead is more
than compensated for by an improved schedule and reduction in number
of facts produced. For MiniMax and N Queens, the number of facts
derived is the same for all configurations, as expected. The
performance seen for those programs arise from reduced memory usage
and improved memory locality.

@@

Fig.~\ref{results:sssp_uspowergrid} shows the number of locks acquired
for the SSSP program.  The first bar in each group shows the number of
locks acquired for the regular version of the program.  The
second bar shows the basic and coordination locks acquired for the coordinated
version of SSSP when every newly derived fact and priority operation is carried out
immediately.  This leads to a significant increase in total locks
acquired often because the priority of a node is change many times
before it is operated on.  The third bar shows the effect of buffering
priority operations before applying them.  This significantly reduces
the lock overhead, but as can be seen in
Table~\ref{queue_instructions}, increases the number of queue operations
slightly.  This slight increase comes from the fact that buffering
delays the change in priorities and some longer paths can sometimes be
investigated unnecessarily.

>>>>

As can be seen in Table~\ref{queue_instructions} all the coordinated
versions of the programs require significantly more queue operations.
The data for this table comes from recording the number of queue
operations and facts derived.  Queue operations represent the number
of normal queue operations executed (each costs 1) plus the number of
\emph{percolate-up}/\emph{percolate-down} operations executed for each
manipulation of the priority queue.  As expected, programs which rely
on prioritizing the order of execution to obtain better asymptotic
complexity (e.g., SSSP), improve in overall running time at the cost of
more queue operations.  In all cases, coordination adds a significant
number of total queue operations, but the resulting overhead is more
than compensated for by an improved schedule and reduction in number
of facts produced. For MiniMax and N Queens, the number of facts
derived is the same for all configurations, as expected. The
performance seen for those programs arise from reduced memory usage
and improved memory locality.

@@

Fig.~\ref{results:sssp_uspowergrid} shows the number of locks acquired
for the SSSP program.  The first bar in each group shows the number of
locks acquired for the regular version of the program.  The
second bar shows the basic and coordination locks acquired for the coordinated
version of SSSP when every newly derived fact and priority operation is carried out
immediately.  This leads to a significant increase in total locks
acquired often because the priority of a node is changed many times
before it is operated on.  The third bar shows the effect of buffering
priority operations before applying them.  This significantly reduces
the lock overhead, but as can be seen in
Table~\ref{queue_instructions}, increases the number of queue operations
slightly.  This slight increase comes from the fact that buffering
delays the change in priorities and some longer paths can sometimes be
investigated unnecessarily.

### 2014-11-13 minimax.tex
Fix verbatim issues.
Fixed several typos.

The code in Fig.~\ref{minimax:check-end} deals with the tree expansion process.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for that create new nodes. In particular, the two
rules in lines 12-26 generate new nodes using the
\mytt{exists} language construct, which creates a child node 
\mytt{B}. We link \mytt{B} with its parent (\mytt{parent(B, A)})
and kick start the expansion of that node \mytt{B} by adding a \mytt{play}
fact.

@@

With coordination, we set the priority of a node to be its depth (lines 15 and
23) so the tree is expanded in a depth-first fashion,
leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads.
Since threads prioritize deeper nodes, the scores of the first leaves are immediately
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

@@

\begin{topfig}
\vspace*{-3ex}
   \begin{center}
      \subfloat[]{\small\begin{tabular}[b]{ | c | c | c |}
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 11.80GB & 0.50MB \\ \hline
         2 & 12.19GB & 1.45MB \\ \hline
         4 & 13.82GB & 2.35MB \\ \hline
         8 & 14.87GB & 4.36MB \\ \hline
         16 & 13.79GB & 8.1MB \\ \hline
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=4.5cm]{results/min-max-tictactoe.png}}
   \end{center}
   \scap{results:memory_minmax}{Memory usage and scalability of the regular and coordinated versions
      of MiniMax.}
\vspace*{-2ex}
\end{topfig}

>>>>

The code in Fig.~\ref{minimax:check-end} deals with the tree expansion process.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases that create new nodes. In particular, the two
rules in lines 12-26 generate new nodes using the
\mytt{exists} language construct, which creates a child node 
\mytt{B}. We link \mytt{B} with its parent (\mytt{parent(B, A)})
and kick start the expansion of that node \mytt{B} by adding a \mytt{play}
fact.

@@

With coordination, we set the priority of a node to be its depth (lines 15 and
23) so that the tree is expanded in a depth-first fashion,
leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads.
Since threads prioritize deeper nodes, the scores of the first leaves are immediately
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

@@

\begin{topfig}
\vspace*{-3ex}
   \begin{center}
      \subfloat[]{\footnotesize\begin{tabular}[b]{ | c | c | c |}
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 11.80GB & 0.50MB \\ \hline
         2 & 12.19GB & 1.45MB \\ \hline
         4 & 13.82GB & 2.35MB \\ \hline
         8 & 14.87GB & 4.36MB \\ \hline
         16 & 13.79GB & 8.1MB \\ \hline
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=4.5cm]{results/min-max-tictactoe.png}}
   \end{center}
   \scap{results:memory_minmax}{Memory usage and scalability of the regular and coordinated versions
      of MiniMax.}
\vspace*{-2ex}
\end{topfig}

### 2014-11-13 nqueens.tex
space saving edits
Fixed several typos.

Since computation goes from the top row to the bottom row, not all
placements of nodes to threads perform equally. This is especially
true because the bottom rows tend to perform the most work. The best
placement is then to split the board vertically so each thread gets
the same number of columns with axiom \mytt{set-cpu(A, vertical(X, Y)}, where
\mytt{X} and \mytt{Y} are the coordinates of the board. Our experiments show that, on a shared
memory system, it does not matter much if we start with a bad
placement since node stealing overcomes those issues by load balancing
dynamically.  However, the N Queens program incrementally builds and shares lists representing
valid board states that are transmitted from top to bottom.
In order to improve memory locality we should then manipulate board
states that share a significant number of elements because each board state
needs to be iterated before extended with a new position.
To accomplish this, we used coordination axioms \mytt{set-default-priority(A,
X)}, where \mytt{X} is the row number (0 is the top row) in order to
prioritize the evaluation of squares closer to the bottom of the board.

@@

\begin{topfig}
   \begin{center}
      \includegraphics[width=\figsize]{results/8queens-13.png}
   \end{center}
   \scap{results:nqueens}{Experimental results for the 13 Queens program.
   Coordination works best when 1 column is assigned to one thread and bottom
   rows have higher priority.}
\end{topfig}

>>>>

Since computation goes from the top row to the bottom row, not all placements of
nodes to threads perform equally. This is especially true because the bottom
rows tend to perform the most work. The best placement is then to split the
board vertically with axiom \mytt{set-cpu(A, vertical(X, Y))} so each thread
gets the same number of columns, where \mytt{X} and \mytt{Y} are the coordinates
of the board. Our experiments show that, on a shared memory system, it does not
matter much if we start with a bad placement since node stealing overcomes those
issues by load balancing dynamically.  However, the N Queens program
incrementally builds and shares lists representing valid board states that are
transmitted from top to bottom.  In order to improve memory locality we should
then manipulate board states that share a significant number of elements because
each board state needs to be iterated before extended with a new position.  To
accomplish this, we used coordination axioms \mytt{set-default-priority(A, X)},
where \mytt{X} is the row number (0 is the top row) in order to
prioritize the evaluation of squares closer to the bottom of the
board.

@@

\begin{topfig}
\vspace*{-1ex}
   \begin{center}
      \includegraphics[width=\figsize]{results/8queens-13.png}
   \end{center}
   \scap{results:nqueens}{Experimental results for the 13 Queens program.
   Coordination works best when 1 column is assigned to one thread and bottom
   rows have higher priority.}
\end{topfig}

### 2014-11-13 sssp.tex
Fix verbatim issues.
Fixed several typos.

When run with one thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. When using multiple
threads, each thread will pick the smallest distance from their subset of nodes.
While this does not yield the optimal program with relation to 1 thread, it
allows for parallel execution and locally avoids unnecessary work. The result
scales well and it is close to Dijkstra's algorithm.

\begin{topfig}
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
type route edge(node, node, int).
type linear shortest(node, int, list int).
type linear relax(node, int, list int).

@@

The most interesting property of the SSSP program presented in
Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
although it applies rules using smarter ordering and the code remains
declarative. The proof of correctness considers that, eventually, the shortest
path is computed at all nodes of the graph until all shortest paths are found.
The derivation of \mytt{set-priority} does not change this
proof.

Fig.~\ref{results:sssp_uspowergrid} shows experimental results when
the program computes the SSSP for 20\% of the nodes of the graph.  The
coordinated version scales well and compared to the regular version
using 16 threads is 1.3 times faster.  the coordinated version
produces 40\% fewer facts when compared to the regular version.
Note that there are some situations where unnecessary facts are
propagated because although the shortest distance is selected locally,
sub-optimal distances may be propagated because many SSSP distances
are computed at the same time.

>>>>

When run with one thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. When using multiple
threads, each thread will pick the shortest distance from their subset of nodes.
While this does not yield the optimal program with relation to 1 thread, it
allows for parallel execution and locally avoids unnecessary work. The result
scales well and it is close to Dijkstra's algorithm.

\begin{topfig}
\scriptsize\begin{Verbatim}[numbers=left,xleftmargin=7mm,commandchars=\\\{\}]
type route edge(node, node, int).
type linear shortest(node, int, list int).
type linear relax(node, int, list int).

@@

The most interesting property of the SSSP program presented in
Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
although it applies rules using smarter ordering and the code remains
declarative. Since the proof of correctness considers that, eventually, the shortest
path is computed at all nodes of the graph, the use of \mytt{set-priority}
does not change this at all.

Fig.~\ref{results:sssp_uspowergrid} shows experimental results when
the program computes the SSSP for 20\% of the nodes of the graph.  The
coordinated version scales well and when compared to the regular version
using 16 threads, it is 1.3 times faster. The coordinated version
also produces 40\% fewer facts with any number of threads.
Note that there are some situations where unnecessary facts are
propagated because although the shortest distance is selected locally,
sub-optimal distances may be propagated because many SSSP distances
are computed at the same time.
